\textbf{Parallel dragonfly algorithm} \\[2ex] \large High Performance Computing for Data Science course 2024/2025

### 2025-08-29 text-paper/paper.tex
paper

\documentclass[10pt]{article}
\usepackage{multicol}
\usepackage{authblk}
\usepackage{mathptmx}
\usepackage{graphicx}
\setlength{\columnsep}{5mm}
\usepackage{blindtext}
\usepackage{geometry}
\usepackage{url}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{float}
\usepackage{longtable}
\usepackage{cite}

@@

\section{Introduction}
Swarm intelligence algorithms are inspired by the collective behavior of biological systems, where simple rules followed by individuals lead to the emergence of complex and adaptive group behavior. The concept was first introduced in Reynolds' Boids model (1987) \cite{TODO}, which simulated flocking behavior, and gained significant traction in Computer Science with the Ant Colony Optimization algorithm (1995) \cite{TODO}.
These algorithms are meta-heuristic in nature, leveraging decentralized and parallel exploration-exploitation strategies to solve optimization problems.

The dragonfly algorithm, first proposed in 2014 and published in 2015 \cite{DAReview}, is a bio-inspired optimization algorithm that simulates the behavior of dragonfly swarms.
By mimicking natural behaviors such as alignment, separation, cohesion, attraction to food sources, and distraction from enemies, the algorithm achieves a balance between exploration and exploitation.
This makes it particularly effective at navigating complex search spaces while reducing the likelihood of being trapped in local optima.
As of 2019, the algorithm has been cited in over 300 scientific papers, highlighting its impact and popularity in the field.

@@

\noindent Notice that the computation of velocities and position updates is compacted into the same \textbf{for} loop.
This optimization reduces execution time by performing all calculations locally within the loop, eliminating the need for auxiliary arrays to store intermediate coefficients.
The rationale for dividing the separation coefficient by the number of neighbors will be explained later in this paper (TODO reference).

@@

\section{Final execution and results}
Now that we have described all the procedure it is time to illustrate how 
the algorithm has been executed. This job has been performed by the
High Performance Computing cluster at University of Trento, which 
use a particular technology called PBS to handle multiple tasks at the same
time: this feature comes useful in our case since, as we will see later, we 
execute multiple istances of the same problem
\subsection*{structure of the execution}

As anticipated we will execute the same problem using multiple instances with different amount of
resources in order to get a complete evaluation of performance. The overall execution use from 1 to 8
threads, and 4 to 1024 cores

@@

After the end of all tasks it is possible to check how much gain in performance we have obtained.
To do this it is necessary to introduce some theoretical concepts that will give us the tools that
will give us a good metric system to evaluate our results.Speedup is defined as the ratio between the the execution times of the serial and parallel
implementations, so the formula will be equal to: $$Speedup = \frac{T_{serial}}{T_{parallel}}$$ 
from this definition it is obvious that the more the speedup is the more the parallel implementation
performs better. There is a particular case where the time execution of the denominator can be defined
as $T_{parallel}=T_{serial}/p$, where \textit{p} is the number of cores: this is called \textit{linear speedup},
and in this situation it perfectly match with the classic speedup coefficient
Efficiency is another important parameter, and it is defined by the ratio of the speedup and the 
number of cores, giving the following formula: $$E = \frac{Speedup}{p}$$
also in this case the efficiency decrease due to the increase of the number of cores. In case
of linear speedup we get that the efficiency is equal to the inverse of $p$
\subsection*{results}
\label{results}
Now it is possible to show results obtained from the cluster,
using the configuration with 4 cores and 1 thread will be considered
equal as the serial time. The reason for this approach is that a single core computation would require 1024 times the workload of the 1024 process version, but we calibrated the 1024 core version to take 3 minutes.

\begin{tabular}{| c | c | c | c | c | c | c |}
  \hline
  T/p & 4	 & 8	& 16	& 32	& 64 & 128  \\
  \hline
  1 & 1	& 2,03 & 3,87 & 7,94 & 14,76 & 31,19  \\
  2 & 1,04 & 2,05 & 3,97 & 8,50 & 16,43 & 27,90 \\	
  4 & 1,01 &2,03	& 4,21 &	8,04 &	16,37 &	35,26	\\	
  8 & 1,02  &2,02 &	4,02 &	8,62 &	18,32 &	32,26 \\
  \hline
\end{tabular}
\newline\newline As we can see the speedup tends to 
increase when the number of cores increase, however when
we use more threads it happens that the
speedup start to decrease at a certain point for some
core configuration. As we can see from the plots it 
seems there are threshold at some specific points, depending
on the number of threads used; notice how the inclinations
of the lines in the plot tends to be less and less evident.

@@

\noindent 
Efficiency tends to present an unpredictable behaviour,
in fact the coefficients grows for some particular setups,
for others instead numbers decrease. If we take a look on
the plot below it shows that there is a high variance in the
datas, with the lines that tends to be more stretched when the
number of threads gets higher
\newline
\newline
\begin{tabular}{| c | c | c | c | c | c | c |}
  \hline
  T/p & 4	 & 8	& 16	& 32	& 64 & 128  \\
  \hline
  1 & 1	& 1,02 & 0,97 & 0,99 & 0,92 & 0,97  \\
  2 & 1,04 & 1,03 & 0,99 & 1,06 & 1,03 & 0,87 \\	
  4 & 1,01	&1,01&	1,05&	1&	1,02&	1,1	\\	
  8 &1,02&	1,01&	1,01&	1,08&	1,15&	1,01 \\
  \hline
\end{tabular}

@@

\end{multicols}

>>>>

\documentclass[10pt]{article}
\usepackage{multicol}
\usepackage{authblk}
\usepackage{mathptmx}
\usepackage{graphicx}
\setlength{\columnsep}{5mm}
\usepackage{blindtext}
\usepackage{geometry}
\usepackage{url}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{float}
\usepackage{longtable}
\usepackage{tabularx}
\usepackage{cite}

@@

\section{Introduction}
Swarm intelligence algorithms are inspired by the collective behavior of biological systems, where simple rules followed by individuals lead to the emergence of complex and adaptive group behavior.
The concept was first introduced in Reynolds' Boids model (1987) \cite{Boids}, which simulated flocking behavior, and gained significant traction in Computer Science with the Ant Colony Optimization algorithm (1995) \cite{Ant_Colony_optimization}.
These algorithms are meta-heuristic in nature, leveraging decentralized and parallel exploration-exploitation strategies to solve optimization problems.

The dragonfly algorithm, first proposed in 2014 and published in 2015 \cite{Original}, is a bio-inspired optimization algorithm that simulates the behavior of dragonfly swarms.
By mimicking natural behaviors such as alignment, separation, cohesion, attraction to food sources, and distraction from enemies, the algorithm achieves a balance between exploration and exploitation.
This makes it particularly effective at navigating complex search spaces while reducing the likelihood of being trapped in local optima.
As of 2019 \cite{Meraihi2020}, the algorithm has been cited in over 300 scientific papers, highlighting its impact and popularity in the field.

@@

\noindent Notice that the computation of velocities and position updates is compacted into the same \textbf{for} loop.
This optimization reduces execution time by performing all calculations locally within the loop, eliminating the need for auxiliary arrays to store intermediate coefficients.
The rationale for dividing the separation coefficient by the number of neighbors will be explained later in this paper in section \ref{chunks}.

@@

\section{Final execution and results}
Now that we have described all the procedure it is time to illustrate how 
the algorithm has been executed. This job has been performed by the
High Performance Computing cluster at University of Trento, which 
use a particular technology called PBS to handle multiple tasks at the same
time: this feature comes useful in our case since, as we will see later, we 
execute multiple istances of the same problem
\subsection*{Structure of the execution}

As anticipated we will execute the same problem using multiple instances with different amount of
resources in order to get a complete evaluation of performance. The overall execution use from 1 to 8
threads, and 4 to 2048 cores. Also we limit the total amount of threads to 2048 for cluster limitation, for a total of 34 runs, runned twice.

@@

After the end of all tasks it is possible to check how much gain in performance we have obtained.
To do this it is necessary to introduce some theoretical concepts that will give us the tools that
will give us a good metric system to evaluate our results.Speedup is defined as the ratio between the the execution times of the serial and parallel
implementations, so the formula will be equal to: $$Speedup = \frac{T_{serial}}{T_{parallel}}$$ 
from this definition it is obvious that the more the speedup is the more the parallel implementation
performs better. There is a particular case where the time execution of the denominator can be defined
as $T_{parallel}=T_{serial}/p$, where \textit{p} is the number of cores: this is called \textit{linear speedup},
and in this situation it perfectly match with the classic speedup coefficient
Efficiency is another important parameter, and it is defined by the ratio of the speedup and the 
number of cores, giving the following formula: $$E = \frac{Speedup}{p}$$
also in this case the efficiency decrease due to the increase of the number of cores. In case
of linear speedup we get that the efficiency is equal to the inverse of $p$
\subsection*{results}
\label{results}
Now it is possible to show results obtained from the cluster,
using the configuration with 4 cores and 1 thread will be considered
equal as the serial time. The reason for this approach is that a single core computation would require 1024 times the workload of the 1024 process version, but we calibrated the 1024 core version to take 3 minutes;
so it should take around $3*1024/4=768min \approx 13h $, computing single core would have taken 4 times that.
\begin{table}[H]
\centering
\tiny  % Use tiny font size for maximum space saving
\setlength{\tabcolsep}{3pt}  % Reduce column separation
\begin{tabular}{|c|*{10}{c}|}
  \hline
  T/P & 4 & 8 & 16 & 32 & 64 & 128 & 256 & 512 & 1024 & 2048 \\
  \hline
  1 & -- & 14260.97 & 8471.77 & 3521.04 & 1849.25 & 1157.88 & 483.40 & 255.68 & 193.13 & 118.08 \\
  2 & -- & 9659.36 & 5628.55 & 3259.06 & 2127.71 & 921.88 & 433.69 & 281.84 & 123.21 & -- \\
  4 & 12785.30 & 7745.36 & 4652.30 & 2398.08 & 1465.56 & 687.89 & 453.28 & 252.71 & -- & -- \\
  8 & 13781.07 & 11326.55 & 3800.54 & 2358.14 & 1308.72 & 722.74 & 597.09 & -- & -- & -- \\
  \hline
\end{tabular}
\end{table}

As we can see the speedup increase almost linearly with the amount of processes used, in the efficency table it is shown that the efficency stays close to 1, until 512 process, then it deeps down. Probably it happens because the distance of the nodes is increasing (maybe throught a slower switch).
Instead increasing thread parallelism is not as effective, and it seems to stagnate after 2 threads (but converging to 2)
The reason could be that mpi pings cpu cores and process executions, so it could be limited to two (becuase Simultaneous Multithreading is usually set to 2).

@@

\noindent 
Efficiency tends to present an unpredictable behaviour,
in fact the coefficients grows for some particular setups,
for others instead numbers decrease. If we take a look on
the plot below it shows that there is a high variance in the
\begin{table}[H]
\centering
\begin{tabular}{| c | c | c | c | c | c | c |}
  \hline
  T/p & 4	 & 8	& 16	& 32	& 64 & 128  \\
  \hline
  1 & 1	& 1,02 & 0,97 & 0,99 & 0,92 & 0,97  \\
  2 & 1,04 & 1,03 & 0,99 & 1,06 & 1,03 & 0,87 \\	
  4 & 1,01	&1,01&	1,05&	1&	1,02&	1,1	\\	
  8 &1,02&	1,01&	1,01&	1,08&	1,15&	1,01 \\
  \hline
\end{tabular}
\caption{Efficiency results}
\label{efficiency_results}
\end{table}

@@

\end{multicols}
\section{Final execution and results}
To sum up, the presented code is highly flexible and scalable, and it could be used for small optimization on low end systems as well as training other methods in an hpc architecture.
The code is available to github, and it can be freely used in different applications.
